Automated Formal Synthesis of Digital Controllers for State-Space Physical Plants

### 2017-01-19 cav2017/cav2017.tex
some revision of sec 2.5 (more needed)

Let $\mathbb{R} \langle I,F \rangle$ be a fixed point domain with $I$ bits representing the integer part and $F$ bits
representing the decimal part. We will extend this notation to $\mathbb{R} \langle I,F,E \rangle$ to add a number of
leading or trailing bits when the domain requires prescaling (ie the maximum is smaller than $0.5$ or the minimum
is greater than $2$)
Let also $\mathbb{R}_* \langle M,E \rangle$ be a floating point domain with $M$ bits
mantissa and $E$ bits exponent. This domain corresponds to 
$$\mathbb{R}_* \langle M,E \rangle = \bigcup_i \mathbb{R} \langle 1,M-1,E_i \rangle : 2^{1-E} \leq E_i  \leq 2^{E-1}$$
This means that the distance between any two numbers in the domain depends on the
location of those numbers.
Since we've shown that a floating point domain is a union of a number of fixed point domains, we will continue our
reasoning on fixed point only. Our points will apply to the floating point domain by extension.
The following approximation errors will arise:
\begin{enumerate}
\item Truncation\\
Let $\mathcal{F}\langle I,F \rangle(\cdot) : \mathbb{R} \rightarrow \mathbb{R}\langle I,F \rangle, \tilde x=\mathcal{F}\langle I,F \rangle(x) = x-\delta : \delta=x\ \%_F\  c\tilde x$, where $\%_F$ is the modulus operation performed on the last bit of the representation ($2^{-F}$).\\
The number $\delta$ corresponds to the truncation error of the representation and it will propagate across operations.
\item Rounding\\
Every time an operation is performed on a number in the $\mathbb{R} \langle I,F,E \rangle$ domain, an error may be
introduced due to truncation of the result.
Let us define $c\langle i,f,e \rangle \in \mathbb{R}\langle I,F,E\rangle$, where $i \in [-2^{I-1}\ 2^{I-1}],f \in [0 1),e \in [-2^{E-1}\ 2^{E-1}]$ are its assigned values. The smallest number that can be represented for any given exponent is $c_{min(e)}=c\langle 0,2^{-F},e\rangle$.
The following errors appear in simple operations:
\begin{enumerate}
\item Addition/Substraction: $c\langle \cdot,e_1\rangle \pm c\langle \cdot,e_2\rangle=c\langle \cdot,max(e_1,e_2)\rangle + \delta_3 : |\delta_3| \leq (e_1==e_2) ? 0 : c_{min(max(e_2,e_1))}$ 
\item Multiplication: $c\langle \cdot,e_1\rangle * c\langle \cdot,e_2\rangle=c\langle \cdot,e_1+e_2\rangle + \delta_3 : |\delta_3| \leq c_{min(e_2+e_1)}$
\end{enumerate}
The bounds obtained for division are left to the reader as it will not be applied in this paper.

At this point it is worth mentioning that not all systems truncate equally. In the definition above $\delta$ may be positive
or negative, and even this decision can be made dependent on the sign of $x$. Common cases in computer systems are:
round downwards (\emph{ie} to $-\infty$), round upwards (\emph{ie} to $+\infty$), and round to zero.
Rounding errors are cumulative which means the overall error will statistically increase with every operation, thus algorithms
performing fewer operations can be more precise in this respect than iterative ones. 
\item Overflow\\
Another source of error introduced by Finite Word Length representations is that of overflow. The finite word length set a
maximum representable number in the domain ($\pm 2^{I-1}+1-2^{-F}$), which means that any number outside this range
cannot be accurately represented by the domain (\emph{eg} $\delta=x-2^{I-1}+1-2^{-F}$). In the case of floating point, a special
case for infinity has been introduced, which addresses some of this issues, but once the error has gone to infinity, the numbers
become meaningless. It is the responsibility of software designers and tools to account for overflows and either supply saturation
functions to prevent them or raise alarms in their presence. In the case of verification, the latter is seen as a safety problem.
\item Null/Zeno behaviour\\
Whilst this work does not explore the effect of Zeno behaviours, it is worth noting that this effect may also be caused by rounding
errors in floating point programs. The reason for this is that if an operation contains two numbers which belong to disjoint sub-domains
(\emph{ie} neither number can be represented in the domain of its counterpart), the resulting operation may become a no-op/skip,
subsequently destabilizing the program.
\end{enumerate}

>>>>

Let $\mathbb{R} \langle I,F \rangle$ be a fixed point domain with $I$ bits representing the integer part and $F$ bits
representing the decimal part. We will extend this notation to $\mathbb{R} \langle I,F,E \rangle$ to add a number of
leading or trailing bits when the domain requires prescaling (ie the maximum is smaller than $0.5$ or the minimum
is greater than $2$).
Let also $\mathbb{R}_* \langle M,E \rangle$ be a floating point domain with $M$ bits
mantissa and $E$ bits exponent. This domain corresponds to 
$$\mathbb{R}_* \langle M,E \rangle = \bigcup_i \mathbb{R} \langle 1,M-1,E_i \rangle : 2^{1-E} \leq E_i  \leq 2^{E-1}$$
This means that the distance between any two numbers in the domain depends on the
location of those numbers.
Since we've shown that a floating point domain is a union of a number of fixed point domains, we will continue our
reasoning on fixed point only. Our reasoning will apply to the floating point domain by extension.
The following approximation errors will arise:
\begin{enumerate}
\item {\bf Truncation:}
Let $\mathcal{F}\langle I,F \rangle(\cdot) : \mathbb{R} \rightarrow \mathbb{R}\langle I,F \rangle, \tilde x=\mathcal{F}\langle I,F \rangle(x) = x-\delta : \delta=x\ \%_F\  c\tilde x$, where $\%_F$ is the modulus operation performed on the last bit of the representation ($2^{-F}$).
The number $\delta$ corresponds to the truncation error of the representation and it will propagate across operations.
\item {\bf Rounding:}
%% Every time an operation is performed on a number in the $\mathbb{R} \langle I,F,E \rangle$ domain, an error may be
%% introduced due to truncation of the result.
Let us define $c\langle i,f,e \rangle \in \mathbb{R}\langle I,F,E\rangle$, where $i \in [-2^{I-1}\ 2^{I-1}],f \in [0 1),e \in [-2^{E-1}\ 2^{E-1}]$ are its assigned values. The smallest number that can be represented for any given exponent is $c_{min(e)}=c\langle 0,2^{-F},e\rangle$.
The following errors appear in simple operations:
\begin{enumerate}
\item Addition/Substraction: $c\langle \cdot,e_1\rangle \pm c\langle \cdot,e_2\rangle=c\langle \cdot,max(e_1,e_2)\rangle + \delta_3 : |\delta_3| \leq (e_1==e_2) ? 0 : c_{min(max(e_2,e_1))}$ 
\item Multiplication: $c\langle \cdot,e_1\rangle * c\langle \cdot,e_2\rangle=c\langle \cdot,e_1+e_2\rangle + \delta_3 : |\delta_3| \leq c_{min(e_2+e_1)}$
\end{enumerate}
The bounds obtained for division are left to the reader as it will not be applied in this paper.

At this point it is worth mentioning that not all systems truncate equally. In the definition above $\delta$ may be positive
or negative, and even this decision can be made dependent on the sign of $x$. Common cases in computer systems are:
round downwards (\emph{ie} to $-\infty$), round upwards (\emph{ie} to $+\infty$), and round to zero.
Rounding errors are cumulative which means the overall error will statistically increase with every operation, thus algorithms
performing fewer operations can be more precise in this respect than iterative ones. 
\item {\bf Overflow:}
  %Another source of error introduced by Finite Word Length representations is that of overflow.
  The finite word length set a
maximum representable number in the domain ($\pm 2^{I-1}+1-2^{-F}$), which means that any number outside this range
cannot be accurately represented by the domain (\emph{eg} $\delta=x-2^{I-1}+1-2^{-F}$). In the case of floating point, a special
case for infinity has been introduced, which addresses some of this issues, but once the error has gone to infinity, the numbers
become meaningless. It is the responsibility of software designers and tools to account for overflows and either supply saturation
functions to prevent them or raise alarms in their presence. In the case of verification, the latter is seen as a safety problem.
\item {\bf Null/Zeno behaviour:}
Whilst this work does not explore the effect of Zeno behaviours, it is worth noting that this effect may also be caused by rounding
errors in floating point programs. The reason for this is that if an operation contains two numbers which belong to disjoint sub-domains
(\emph{ie} neither number can be represented in the domain of its counterpart), the resulting operation may become a no-op/skip,
subsequently destabilizing the program.
\end{enumerate}
